---
title: Interactive COVID-19 maps by piggybacking (Part 1)
author: me
date: '2020-10-13'
slug: testing-static-files
categories:
  - Development
tags:
  - R Markdown, Kaggle, Plotly, John Hopkins University
description: ''
topics: [interactive visuals, interactive maps, COVID-19]
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE)
```

The COVID-19 pandemic cast a spotlight on the need for reliable data and responsible analysis to track and minimize the loss of human life. 
Among the many datasets and reports made available in recent months, 
[this author's work caught my attention](https://www.kaggle.com/gerardopgarcia/coronavirus-covid-19-visualization-prediction/){target="_blank"}, compiling large amounts of data into one report that was easy to maintain. While their work
is thorough and, in my opinion, well-written, it's difficult to digest due to its volume. I also recently came across [an example using geographic
data to make interactive maps using R and Plotly (Javascript)](https://plotly.com/r/sliders/){target="_blank"} 
and thought: what if I combined these to make a single visual that condensed pages of information?

The data are provided by John Hopkins University and updated frequently (if not daily), 
[available to the public via Github](https://github.com/CSSEGISandData/COVID-19){target="_blank"}. Their data dashboard 
[can be found here](https://www.arcgis.com/apps/opsdashboard/index.html#/bda7594740fd40299423467b48e9ecf6){target="_blank"}. While the dashboard already provides multiple 
visuals, this small project allows us to choose and download different 3-D map projections of the 
Earth (or U.S.), as well as change the map's focus by either clicking and dragging manually or using sliders to adjust the geographic coordinates at the center.

The data are available as CSV files through their respective hyperlinks, and, once downloaded, I changed all the column names to lowercase for reference. The majority are time-series data that include coordinates of multiple COVID-19 measures for countries 
and sub-regions globally or specific to the U.S. The use of hyperlinks also ensure that the most up-to-date is always used. 
For now, I focus on the latest <em>confirmed</em> COVID-19 cases only for simplicity.

<details><summary><em>Click here to show code.</em></summary>

```{r}
# Gather data from URLs in CSV format

confirmed_df <- read.csv(
  'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv')

deaths_df <- read.csv(
  'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv')

recoveries_df <- read.csv(
  'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv')

latest_data <- read.csv(
  'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-11-2020.csv')

us_medical_data <- read.csv(
  'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports_us/10-11-2020.csv')

apple_mobility <- read.csv(
  'https://covid19-static.cdn-apple.com/covid19-mobility-data/2018HotfixDev24/v3/en-us/applemobilitytrends-2020-10-11.csv')

# Change columns to lower case

names(latest_data) <- tolower(names(latest_data))

names(confirmed_df) <- tolower(names(confirmed_df))

names(us_medical_data) <- tolower(names(us_medical_data))

names(deaths_df) <- tolower(names(deaths_df))

names(recoveries_df) <- tolower(names(recoveries_df))

names(apple_mobility) <- tolower(names(apple_mobility))
```

</details>

<hr>

<b>Latest data from John Hopkins University (<em>requested `r Sys.Date()`</em>):</b>

<div style="overflow-x:scroll;">

```{r echo=FALSE}
DT::datatable(latest_data,
  filter='top',
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip',
  buttons = c('copy', 'csv', 'excel', 'pdf'),
  pageLength=10,
  lengthMenu = list(c(10,25,50,-1),
  c(10,25,50,"All"))),
  rownames = FALSE)
```

</div>

I then turned to the Plotly example I found online, which rendered a 3-D globe and drop-down menu of numerous map projects to choose from.
The data and code used to draw the lines were identified and eventually removed altogether, leaving what is essentially an empty 
container to use as a template.

<hr>

<b>Original example of interactive maps from Plotly:</b>

<details><summary><em>Click here to show code.</em></summary>

```{r}
library(plotly)

df <- read.csv('https://raw.githubusercontent.com/plotly/datasets/master/globe_contours.csv')
df$id <- seq_len(nrow(df))

library(tidyr)
d <- df %>%
  gather(key, value, -id) %>%
  separate(key, c("l", "line"), "\\.") %>%
  spread(l, value)

geo <- list(
  showland = TRUE,
  showlakes = TRUE,
  showcountries = TRUE,
  showocean = TRUE,
  countrywidth = 0.5,
  landcolor = 'rgb(230, 145, 56)',
  lakecolor = 'rgb(0, 255, 255)',
  oceancolor = 'rgb(0, 255, 255)',
  projection = list(
    type = 'orthographic',
    rotation = list(
      lon = -100,
      lat = 40,
      roll = 0
    )
  ),
  lonaxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  ),
  lataxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  )
)

## add custom events

# dropdown
projections = data.frame(type = c(
  "equirectangular", "mercator", "orthographic", "natural earth","kavrayskiy7", 
  "miller", "robinson", "eckert4", "azimuthal equal area","azimuthal equidistant", 
  "conic equal area", "conic conformal", "conic equidistant", "gnomonic", "stereographic", 
  "mollweide", "hammer", "transverse mercator", "albers usa", "winkel tripel"))

all_buttons <- list()
for (i in 1:length(projections[,])) { 
  all_buttons[[i]] <- list(method = "relayout",
  args = list(list(geo.projection.type = projections$type[i])),
  label = projections$type[i])
}

# sliders
lon_range = data.frame(seq(-180, 180, 10))
lat_range = data.frame(seq(-90, 90, 10))
colnames(lon_range) <- "x"
colnames(lat_range) <- "x"

all_lat <- list()
for (i in 1:length(lat_range[,])) { 
  all_lat[[i]] <- list(method = "relayout",
                       args = list(list(geo.projection.rotation.lat = lat_range$x[i])),
                       label = lat_range$x[i])
}

all_lon <- list()
for (i in 1:length(lon_range[,])) {  
  all_lon[[i]] <- list(method = "relayout", 
  args = list(list(geo.projection.rotation.lon = lon_range$x[i])),
  label = lon_range$x[i]) 
} 

# annotations
annot <- list(x = 0, y=0.8, text = "Projection", yanchor = 'bottom', 
  xref = 'paper', xanchor = 'right',
  showarrow = FALSE)


# original d3-globe with contours
fig<- plot_geo(d) 
fig <- fig %>% group_by(line) 
fig <- fig %>% add_lines(x = ~lon, y = ~lat, color = ~line, colors = 'Reds') 
fig <- fig %>% layout(
  showlegend = FALSE, geo = geo
)

# plot with custom events
fig<- fig
fig <- fig %>% layout(annotations = annot,
  updatemenus = list(list(active = 2, x = 0, y = 0.8, 
    buttons=all_buttons)),
  sliders = list(
    list(
      active = (length(lon_range[,])-1)/2, 
      currentvalue = list(prefix = "Longitude: "), 
      pad = list(t = 20), 
      steps = all_lon),
    list(
      active = (length(lat_range[,])-1)/2, 
      currentvalue = list(prefix = "Latitude: "), 
      pad = list(t = 100), 
      steps = all_lat)))
```

</details>

```{r echo=FALSE}
fig
```

<hr>

<b>Empty 'container' remaining after removing irrelevant code:</b>

<details><summary><em>Click here to show code.</em></summary>

```{r}
library(plotly)

# REMOVED
# ------------------------------------------------------------------------------
# df <- read.csv('https://raw.githubusercontent.com/plotly/datasets/master/globe_contours.csv')
# df$id <- seq_len(nrow(df))
# 
# library(tidyr)
# d <- df %>%
#   gather(key, value, -id) %>%
#   separate(key, c("l", "line"), "\\.") %>%
#   spread(l, value)
# ------------------------------------------------------------------------------

geo <- list(
  showland = TRUE,
  showlakes = TRUE,
  showcountries = TRUE,
  showocean = TRUE,
  countrywidth = 0.5,
  landcolor = 'rgb(230, 145, 56)',
  lakecolor = 'rgb(0, 255, 255)',
  oceancolor = 'rgb(0, 255, 255)',
  projection = list(
    type = 'orthographic',
    rotation = list(
      lon = -100,
      lat = 40,
      roll = 0
    )
  ),
  lonaxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  ),
  lataxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  )
)

## add custom events

# dropdown
projections = data.frame(type = c(
  "equirectangular", "mercator", "orthographic", "natural earth","kavrayskiy7", 
  "miller", "robinson", "eckert4", "azimuthal equal area","azimuthal equidistant", 
  "conic equal area", "conic conformal", "conic equidistant", "gnomonic", "stereographic", 
  "mollweide", "hammer", "transverse mercator", "albers usa", "winkel tripel"))

all_buttons <- list()
for (i in 1:length(projections[,])) { 
  all_buttons[[i]] <- list(method = "relayout",
  args = list(list(geo.projection.type = projections$type[i])),
  label = projections$type[i])
}

# sliders
lon_range = data.frame(seq(-180, 180, 10))
lat_range = data.frame(seq(-90, 90, 10))
colnames(lon_range) <- "x"
colnames(lat_range) <- "x"

all_lat <- list()
for (i in 1:length(lat_range[,])) { 
  all_lat[[i]] <- list(method = "relayout",
  args = list(list(geo.projection.rotation.lat = lat_range$x[i])),
  label = lat_range$x[i])
}

all_lon <- list()
for (i in 1:length(lon_range[,])) {  
  all_lon[[i]] <- list(method = "relayout", 
  args = list(list(geo.projection.rotation.lon = lon_range$x[i])),
  label = lon_range$x[i]) 
} 

# annotations
annot <- list(x = 0, y=0.8, text = "Projection", yanchor = 'bottom', 
  xref = 'paper', xanchor = 'right',
  showarrow = FALSE)

# REMOVED
# ------------------------------------------------------------------------------
# original d3-globe with contours
# fig<- plot_geo(d) 
# fig <- fig %>% group_by(line) 
# fig <- fig %>% add_lines(x = ~lon, y = ~lat, color = ~line, colors = 'Reds') 
# fig <- fig %>% layout(
#   showlegend = FALSE, geo = geo
# )
# ------------------------------------------------------------------------------

# REPLACED
# ------------------------------------------------------------------------------
fig <- plot_geo()
fig <- fig %>% layout(
  showlegend=FALSE, geo=geo
)
# ------------------------------------------------------------------------------

# plot with custom events
fig<- fig
fig <- fig %>% layout(annotations = annot,
  updatemenus = list(list(active = 2, x = 0, y = 0.8, 
    buttons=all_buttons)),
  sliders = list(
    list(
      active = (length(lon_range[,])-1)/2, 
      currentvalue = list(prefix = "Longitude: "), 
      pad = list(t = 20), 
      steps = all_lon),
    list(
      active = (length(lat_range[,])-1)/2, 
      currentvalue = list(prefix = "Latitude: "), 
      pad = list(t = 100), 
      steps = all_lat)))
```

</details>

```{r echo=FALSE}
fig
```

With the visual essentially an empty container, I reshaped the latest COVID-19 data to fit the strucutre of the original dataset. 
First, I split up the data into 2 groups: those with and without confirmed case counts reported for geographic areas smaller than the country itself. 
Countries that only reported a single statistic for their national as whole were grouped under the color of reds and the others in blues. To create more variation in 
the markers on the map, I sorted the data into decreasing order and assigned them a color by their global rank.

<details><summary><em>Click here to show code.</em></summary>

```{r}
# Restructure latest COVID-19 data for plot_geo visual

# Country-only confirmed case reports

latest_data_countries <- latest_data[latest_data$province_state=="",]

latest_data_countries$rank[order(latest_data_countries$confirmed, decreasing = FALSE)] <- 1:nrow(latest_data_countries)

# All other confirmed case reports

latest_data_sub <- latest_data[!latest_data$province_state=="",]

latest_data_sub$rank[order(latest_data_sub$confirmed, decreasing = FALSE)] <- 1:nrow(latest_data_sub)
```

</details>

And replaced the data into the now empty container. I also included a few options restricting user options to interact with the visual, preventing 
awkward results that could obscure parts of the map until it was reloaded. I included the option to download the current map as a PNG file while removing the Plotly logo 
for stylistic reasons.

<hr>

<b>Final result after merging the new data and empty container:</b>

<details><summary><em>Click here to show code.</em></summary>

```{r}
library(plotly)

library(tidyr)

geo <- list(
  showland = TRUE,
  showlakes = TRUE,
  showcountries = TRUE,
  showocean = TRUE,
  countrywidth = 0.5,
  landcolor = 'rgb(230, 145, 56)',
  lakecolor = 'rgb(0, 255, 255)',
  oceancolor = 'rgb(0, 255, 255)',
  projection = list(
    type = 'orthographic',
    rotation = list(
      lon = -100,
      lat = 40,
      roll = 0
    )
  ),
  lonaxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  ),
  lataxis = list(
    showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  )
)

## add custom events

# dropdown
projections = data.frame(type = c("equirectangular", "mercator", "orthographic", "natural earth","kavrayskiy7", 
  "miller", "robinson", "eckert4", "azimuthal equal area","azimuthal equidistant", 
  "conic equal area", "conic conformal", "conic equidistant", "gnomonic", "stereographic", 
  "mollweide", "hammer", "transverse mercator", "albers usa", "winkel tripel"))

all_buttons <- list()
for (i in 1:length(projections[,])) { 
  all_buttons[[i]] <- list(method = "relayout",
  args = list(list(geo.projection.type = projections$type[i])),
  label = projections$type[i])
}

# sliders
lon_range = data.frame(seq(-180, 180, 10))
lat_range = data.frame(seq(-90, 90, 10))
colnames(lon_range) <- "x"
colnames(lat_range) <- "x"

all_lat <- list()
for (i in 1:length(lat_range[,])) { 
  all_lat[[i]] <- list(method = "relayout",
  args = list(list(geo.projection.rotation.lat = lat_range$x[i])),
  label = lat_range$x[i])
}

all_lon <- list()
for (i in 1:length(lon_range[,])) {  
  all_lon[[i]] <- list(method = "relayout", 
  args = list(list(geo.projection.rotation.lon = lon_range$x[i])),
  label = lon_range$x[i]) 
} 

# annotations

annot <- list(x = 0, y=0.8, text = "<b>Confirmed Cases</b><br><br><br>Projection",
  yanchor = 'bottom', 
  xref = 'paper', xanchor = 'right',
  showarrow = FALSE)

# NEW DATA WAS PLACED HERE
# ------------------------------------------------------------------------------
# COVID-19 data

fig <- plot_geo()

# Add data reported within countries

fig <- fig %>% add_markers(x=~latest_data_sub$long_, y=~latest_data_sub$lat, 
  color=~latest_data_sub$rank, colors = "Blues",
  hovertemplate=~paste("<b><br>",latest_data_sub$country_region, "<br>", 
    latest_data_sub$province_state, "<br>", 
    latest_data_sub$admin2, "<br>", latest_data_sub$confirmed, "</b><extra></extra>"))

# Add data from cases reported for countries in aggregate

fig <- fig %>% add_markers(x=~latest_data_countries$long_, y=~latest_data_countries$lat, 
  color=~latest_data_countries$rank, colors = "Reds",
  hovertemplate=~paste("<b><br>",latest_data_countries$country_region, "<br>", 
    latest_data_countries$confirmed, "</b><extra></extra>"))

# Hide elements of the visual, including legends and gradients

fig <- fig %>% layout(
  showlegend = FALSE, geo = geo
)%>%hide_colorbar() 
# ------------------------------------------------------------------------------

# plot with custom events

fig<- fig
fig <- fig %>% layout(annotations = annot,
  updatemenus = list(list(active = 2, x = 0, y = 0.8, 
  buttons=all_buttons)),
  sliders = list(
    list(
      active = (length(lon_range[,])-1)/2, 
      currentvalue = list(prefix = "Longitude: "), 
      pad = list(t = 20), 
      steps = all_lon),
    list(
      active = (length(lat_range[,])-1)/2, 
      currentvalue = list(prefix = "Latitude: "), 
      pad = list(t = 100), 
      steps = all_lat)))

# Adjust Plotly options

fig <- fig %>%
  config(displaylogo = FALSE)%>%
  config(modeBarButtonsToRemove = c(
    "zoomIn2d", "zoomOut2d", "sendDataToCloud", "editInChartStudio", "zoom2d", 
    "pan2d", "select2d", "lasso2d", "drawclosedpath", "drawopenpath", "drawline", 
    "drawrect", "drawcircle", "eraseshape", "autoScale2d", "resetScale2d", 
    "hoverCloestCartesian", "hoverCompareCartesian", "handleCartesian", 
    "toggleSpikelines"))

# Add title - Not included for now because it cuts off along the top
# Included under the legend annotation for better use of space

# fig <- fig %>%
#  layout(title=list(text='<b>Confirmed COVID-19 Cases</b>'))
```

</details>

```{r echo=FALSE}
fig
```

These steps can be repeated to create interactive maps of active cases, deaths, and recoveries with the available datat. 
The next step will be pulling time-series data and map changes over time. There are tools available through HTML widgets that 
can facilitate this by adding a <em>play</em> button to the map.
