---
title: "Gun Violence"
author: "me"
date: 2020-08-11T21:13:14-05:00
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

Following the outbreak of COVID-19, the Black Lives Matter movement brought attention to police violence against the Black community and the people overall, with protests spanning the globe bringing additional scrutiny and evidence against long-established practices. A wealth of information has followed, including self-recorded videos and data, allowing us to connect systemic-level violence with the intensity of individual shootings that some may have dismissed as anecdotal. The police fatal-shooting data are from [Ahsen Nazir's data set uploaded to Kaggle, downloaded on 8/06/2020](https://www.kaggle.com/ahsen1330/us-police-shootings). The unedited data are accessible in full below, and are offered for transparency and access.

To test static files, try [test](/html/state_reporting.html).


<details><summary>Click to see set up code</summary>
<p>
```{r setup}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

```{r}
# Check and install required libraries.
list.of.packages <- c(
    "dplyr", "ggplot2", "tidyverse", "ggpubr","plotly", "DT")
new.packages <- list.of.packages[!(list.of.packages %in%
    installed.packages()[,"Package"])]

# Remove checked vectors from memory.
if(length(new.packages)) install.packages(new.packages)
rm(list.of.packages, new.packages)
```
</p>
</details>

<details><summary>Click to see custom functions code</summary>
<p>
```{r}
# Functions for Visualizations

# Paste percent symbol
label_percent <- function(x){
  x <- as.character(x)
  for(i in 1:length(x)){
    if(i==1){
      x[i] <- paste(as.numeric(x[i])*100, "%", sep="")
    } else{ 
      x[i] <- paste(as.numeric(x[i])*100, "%", sep="")
    }
  }
  return(x)
}

label_bold <- function(x){
  x <- as.character(x)
  for(i in 1:length(x)){
    if(i==1){
      x[i] <- paste("<b>", x[i], "</b>", sep="")
    } else{ 
      x[i] <- paste("<b>", x[i], "</b>", sep="")
    }
  }
  return(x)
}

# Tables
create_dt <- function(x){
  require(DT)
  DT::datatable(x,
    extensions = 'Buttons',
    options = list(dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel'),
      pageLength=10,
      scrollX=TRUE,
      lengthMenu = list(c(10,25,50,-1),
        c(10,25,50,"All"))),
    rownames = FALSE)
}

# Donut charts
render_donut <- function(df, title_, pct=FALSE){
  require(dplyr)
  require(ggplot2)
  require(plotly)
  require(scales)
  library(stats)

  df <- data.frame(table(df$group))
  df <- df[order(df$Freq, decreasing=T),]
  df <- rename(df, group=Var1, value=Freq)
  row.names(df) <- NULL
    
  
  # Data expect 'group' and 'value' columns...
  
  # tooltip_text <- paste0(
  #   "<b>", df$group, ": </b>", df$value)
  #   if(pct){
  #     tooltip_text <- paste0(
  #       "<b>", df$group, ": </b>", label_percent(df$value))
  #   }
  
  # Define plotly object
 # df <- df %>% group_by(group)
 # df <- df %>% summarize(count = n())
  fig <- df %>% plot_ly(labels = ~group, values = ~value, textinfo= ~value)
  fig <- fig %>% add_pie(hole = 0.6)
  fig <- fig %>% layout(title = paste("<b>", title_, "</b>", sep=""),  showlegend = T,
  xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
  yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))%>%
  config(displaylogo = FALSE)%>%
  config(modeBarButtonsToRemove = c(
  "zoomIn2d", "zoomOut2d", "sendDataToCloud", "editInChartStudio", "zoom2d", 
  "pan2d", "select2d", "lasso2d", "drawclosedpath", "drawopenpath", "drawline", 
  "drawrect", "drawcircle", "eraseshape", "autoScale2d", "resetScale2d", 
  "hoverCloestCartesian", "hoverCompareCartesian", "handleCartesian", 
  "toggleSpikelines"))%>%
  layout(autosize=F, margin=list(t=75, pad=0), #width=400, height=400,
    legend=list(orientation='v', x=1.2, y=0.5, face='bold', font=list(size=15)))

    return(fig)
}

# Create visuals
render_visual <- function(df, title_, pct=FALSE){
  require(dplyr)
  require(ggplot2)
  require(plotly)
  require(scales)
  require(stats)
  
  # # Global constants
  #   #fig_width <- as.numeric(825)
  # #  fig_height <- as.numeric(525)
  #   
  #   # Fix dataframe (top 8)
  #   df <- rename(df, group=response)
  #   
  #   if(nrow(df)>8){
  #       temp <- df[1:8,]
  #       
  #       if(df$value[8]==df$value[9] & df$value[8]!=1){
  #             boundary <- df$value[8] + 1
  #             temp <- df[df$value > boundary,]
  #       }
  #       df <- temp
  #   }
  
  # Define ggplot object
  tooltip_text <- paste0(
    "<b>", df$group, ": </b>", label_percent(df$value), "</br></br><b>Total: </b>", df$n)
    
 
  visual <- df %>% 
    dplyr::mutate(group=factor(label_bold(as.character(group)),
      levels=unique(label_bold(group))))%>%
    ggplot(aes(x=factor(group, levels=rev(levels(group))), y=value, text=tooltip_text))+
      geom_col(position="dodge", color="black", aes(fill=type), alpha=1)+
      coord_flip()+
      labs(title=title_, 
        subtitle = "", 
          x="", y="", fill="")+
      #scale_y_continuous(expand=c(0,0), limits = (0, 100), labels = scales::percent(x=value))+
    scale_y_continuous(expand=c(0,0), labels = percent, limits=c(0,1))+
 #   limits=c(0, 100)+
          #                  
          #                  number_format(accuracy = 1),
          #  # breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1)))))+
          # limits = c(0, (df$value%>%
          # sort(., decreasing = TRUE)%>%
          #   100))) +
          scale_fill_brewer(direction = -1)+
          theme_bw()+
          theme(legend.position = "bottom",
            legend.box="horizontal",
            axis.text.x = element_text(color="black", size = 11),
            axis.text.y = element_text(color="black", size = 11))
                
    visual <- ggplotly(visual, tooltip="text")%>% 
      layout(xaxis=list(fixedrange=TRUE))%>%
      layout(yaxis=list(fixedrange=TRUE))%>%
      layout(title=list(text=paste0(
        '<b>', title_,'</b>')))%>%
        layout(annotations=list(x = 1, y = -0.1, text = "", 
        showarrow = FALSE, xref='paper', yref='paper', 
        xanchor='right', yanchor='auto', xshift=0, yshift=-10,
        font=list(size=12, color="black")))%>%
      layout(legend=list(orientation="h", x=0.25, y=-0.2))%>%
      config(displaylogo = FALSE)%>%
      config(modeBarButtonsToRemove = c(
        "zoomIn2d", "zoomOut2d", "sendDataToCloud", "editInChartStudio", "zoom2d", 
        "pan2d", "select2d", "lasso2d", "drawclosedpath", "drawopenpath", "drawline", 
        "drawrect", "drawcircle", "eraseshape", "autoScale2d", "resetScale2d", 
        "hoverCloestCartesian", "hoverCompareCartesian", "handleCartesian", 
        "toggleSpikelines"))%>%
      layout(margin=0,0,0,0)%>%
      style(textposition="right")
      
  return(visual)}
```

</p>
</details>
```{r}
# Import data
df <- read.csv(
  "C:/Users/Shadow/Documents/Data/Projects/Kaggle/shootings_2020_08_06.csv")

create_dt(df)
```

From the data, we see that men accounted for over 95% of reported fatalities between 2015 and 2020.

```{r}
library(dplyr)

temp <- subset(df, select=c(gender))
temp <- rename(temp, group=gender)

render_donut(temp, 'Fatal Police Shootings by Gender')
rm(temp)
```


Just over 50% of victims were white, [despite accounting for 75% of the US population](https://www.census.gov/quickfacts/fact/table/US/PST045219). Meanwhile, 27% of victims were Black (13% of the population) and 18% were Hispanic (19%).

```{r}
library(dplyr)

temp <- subset(df, select=c(race))
temp <- rename(temp, group=race)

render_donut(temp, 'Fatal Police Shootings by Race')
rm(temp)
```

Using the 2019 US Census to *estimate* the population sizes for each race in the last 5 years, we can begin to understand the racial disparities despite the number of white victims. This assumes racial demographics have remained constant for the last 5 years in the US.

```{r}
library(dplyr)

temp <- subset(df, select=c(race))
temp <- rename(temp, group=race)

# Turn counts into percentages 
temp <- data.frame(table(temp))
temp <- rename(temp, group=temp, value=Freq)
temp$value <- round(temp$value/sum(temp$value), digits=2)
temp <- temp[order(temp$value, decreasing=T),]

# Enter census data (link provided)
group <- c('White', 'Black', 'Hispanic', 'Asian', 'Native', 'Other')

# Race-size percentages in 2019
value <- c(
  0.763,
  0.134,
  0.185,
  0.059,
  0.002,
  0.028)

# Total US population in 2019
total <- 328239523

n <- c(
  round(0.763*total, digits=0),
  round(0.134*total, digits=0),
  round(0.185*total, digits=0),
  round(0.059*total, digits=0),
  round(0.002*total, digits=0),
  round(0.028*total, digits=0)
)

# Total population subset
temp1 <- data.frame(cbind(group, value, n))
temp1$type <- 'Total Population'

# Race subset
temp$n <- round(temp$value*nrow(df), digits=0)
temp$type <- 'Killed by Police'

# Merge & create figure
temp <- rbind(temp, temp1)
temp$value <- as.numeric(temp$value)
render_visual(temp, 
  'Fatalities vs. Race Size from 2015 to 2020\n(Estimated with 2019 US Census)')
rm(temp)
rm(temp1)
rm(group)
rm(n)
rm(value)
rm(total)
```


